##Reproducible Research
#Peer Assessment 1
#2014-05-19

#*Loading and preprocessing the data*
- Load the data
- Process/transform the data
```{r}
## Dependencies

## Set file locations
setwd('~/Documents/repoducibleresearch/')

## Read data
stepdat <-read.csv(unz('repdata-data-activity.zip','activity.csv'))
```

#*What is mean total number of steps taken per day?*
- Histogram of total number of steps taken each day
- Mean and median total number of steps taken each day
```{r}
## Aggregate by day and display histogram
stepsperday <- aggregate(stepdat$steps,by=list(as.numeric(as.Date(stepdat$date))),FUN=sum,na.rm=TRUE)
hist(stepsperday[,2],main='Histogram of Total Number of Steps Per Day', ylab='Frequency',xlab='Steps per Day')

## Mean and median
stepsum <- summary(stepsperday[,2])
```
The mean and median number of steps per day are `r stepsum[3]` and `r stepsum[4]` respectively.

#* What is the average daily activity pattern?*
- Time series plot
- Which 5 minute interval contains maximum # of steps?

```{r, warning=FALSE}
aggdat <- aggregate(stepdat, by=list(stepdat$interval),FUN=mean,na.rm=TRUE)
plot(x=aggdat$interval,y=aggdat$steps,type='l')
maxint <- aggdat[which(aggdat$steps==max(aggdat$steps)),]$interval
```

The interval containing, on average, the maximum numer of steps is `r maxint`.

#*Imputing missing values*
- Calculate and report total number of rows with missing values
- Devise imputation strategy
- Create new data set with missing values imputed
- Histogram of total steps per day, calculate mean and median steps per day

```{r}
missrows <- sum(is.na(stepdat))
```

A total of `r missrows` rows contain missing values.

*Imputation strategy*

Replace missing values with observed mean for that interval. A more sophisticated approach can be found in package mtsdi.

Code below replaces missing values.

```{r,warning=FALSE}
stepdat2 <- stepdat
toreplace <- is.na(stepdat2)
for (i in 1:dim(toreplace)[1]) {
 if (toreplace==TRUE) {
     stepdat2$steps[which(toreplace)] <- aggdat$steps[which(aggdat$interval==stepdat2$interval[which(toreplace)])]
 }
}

## Aggregate imputed data by day
stepsperday2 <- aggregate(stepdat2$steps,by=list(as.numeric(as.Date(stepdat2$date))),FUN=sum,na.rm=TRUE)
hist(stepsperday2[,2],main='Histogram of Imputed Number of Steps Per Day', ylab='Frequency',xlab='Steps per Day')

## Mean and median
stepsum2 <- summary(stepsperday2[,2])
```

The mean and median number of steps per day are `r stepsum2[3]` and `r stepsum2[4]` respectively.

#*Weekday/Weekend Differences*
```{r}
stepdat2$weekday <- as.factor(weekdays(as.Date(stepdat2$date))=='Saturday' | weekdays(as.Date(stepdat2$date))=='Sunday')

```